### 2011-06-10 thesis/novelty.tex
Lots of content and changes on detecting novelty detection without assumption of P(novel) is constant on any structure.

\section{Detecting Novelty on a Single Variable}
This section focus on detecting a novel class on single variable of the
conceptual map. As seen before on \autoref{chap:novelty-intro},
assuming that $P(\overline{novel})$ is constant, novelty detection can be
done by implementing a threshold on $P(x|\overline{novel})/P(x)$.
In order to correctly measure novelty on a single variable both
conditional and unconditional probabilities need to be coherently defined:

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent] {$a$}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \node (shape1f) [factor] {}; \&
    \node (appearance1f) [factor] {}; \&
    \node (object1f) [factor] {}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

@@

\begin{figure}[h]
\centering
\subfloat[\label{fig:novel-a-correct}Graphical model able to correctly represent
          the unconditional probability when variable $a$ can be novel.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [factor] {}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \&
    \&
    \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

@@

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\qquad
\subfloat[By using independent features for modelling the unconditional
          distribution an highly simplified graph is obtained.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \&
    \&
    \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,dashed] {$a^*$}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \nofactor {shape1f}{}{$U$} {right=0pt}; \&
    \nofactor {appearance1f}{} {$U$} {right=0pt}; \&
    \nofactor {object1f}{} {$U$} {right=0pt}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };
  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);
  \node (captr12f) [above=0pt of r12f] {\footnotesize{$U$}};
  \node (captr13f) [above=0pt of r13f] {\footnotesize{$U$}};

@@

Concluding, after merging graph $G$ and $U_a$ into graph $B$, the condition
described in \autoref{eq:threshold-with-ta} can be test by performing a
\gls{MAP} operation on variable $c$ of graph $B$ when the threshold factor
$\phi_t$ is proportional to:
\begin{equation}
\phi_t(c) \propto \left\{
  \begin{array}{l l}
    1, & \qquad \text{if } c = 0 \\
    t_A \frac{Z_G}{Z_{U_a}}, & \qquad \text{if } c = 1\\
  \end{array} \right.
\end{equation}

@@

\begin{figure}[h]
\centering
lalla
\caption{\label{fig:merging-graphs}A condition based on the ratio between 2
         factor graphs that share some structure can be performed by testing the
         MAP probability of a variable on a merged graph.}
\end{figure}

@@

\section{Detecting and Copping with Novelty on Multiple Variable}

>>>>

\section{Detecting Novelty with Ratio Between Two Graphs}
This section focus on detecting a novel class on single variable of the
conceptual map. As seen before on \autoref{chap:novelty-intro},
assuming that $P(\overline{novel})$ is constant, novelty detection can be
done by implementing a threshold on $P(x|\overline{novel})/P(x)$.
In order to correctly measure novelty on a single variable both
conditional and unconditional probabilities need to be coherently defined:

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent] {$a$}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \node (shape1f) [factor] {}; \&
    \node (appearance1f) [factor] {}; \&
    \node (object1f) [factor] {}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

@@

\begin{figure}[h]
\centering
\subfloat[\label{fig:novel-a-correct}Graphical model able to correctly represent
          the unconditional probability when variable $a$ can be novel.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [factor] {}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \&
    \&
    \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

@@

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\qquad
\subfloat[By using independent features for modelling the unconditional
          distribution an highly simplified graph is obtained.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \&
    \&
    \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,dashed] {$a^*$}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \nofactor {shape1f}{}{$U$} {right=0pt}; \&
    \nofactor {appearance1f}{} {$U$} {right=0pt}; \&
    \nofactor {object1f}{} {$U$} {right=0pt}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };
  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);
  \node (captr12f) [above=0pt of r12f] {\footnotesize{$U$}};
  \node (captr13f) [above=0pt of r13f] {\footnotesize{$U$}};

@@

Concluding, after merging graph $G$ and $U_a$ into graph $B$, the condition
described in \autoref{eq:threshold-with-ta} can be test by performing a
\gls{MAP} operation on variable $c$ of graph $B$ when the threshold factor
$\phi_t$ is proportional to:
\begin{equation}
\label{eq:structure-ratio}
\phi_t(c) \propto \left\{
  \begin{array}{l l}
    1, & \qquad \text{if } c = 0 \\
    t_A \frac{Z_G}{Z_{U_a}}, & \qquad \text{if } c = 1\\
  \end{array} \right.
\end{equation}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \node (thresholdguide) [latent,draw=none] {}; \&
    \&
    \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \node (control) [latent,dashed] {$c$}; \&
    \node (room1) [latent,dashed] {$b$}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \nofactor {shape1f}{}{$\phi_b$} {right=0pt}; \&
    \nofactor {appearance1f}{} {$\phi_b$} {right=0pt}; \&
    \nofactor {object1f}{} {$\phi_b$} {right=0pt}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };
  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (control) node (controlf) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);
  \node (captr12f) [above=0pt of r12f] {\footnotesize{$\phi_b$}};
  \node (captr13f) [above=0pt of r13f] {\footnotesize{$\phi_b$}};

  \draw [-,draw=none] (room1) -- (control) node (controlf) [midway,factor] {};
  \draw [-,draw=none] (control) -- (thresholdguide) node (thresholdf) [midway,factor] {};
  \node (captcontrolf) [above=0pt of controlf] {\footnotesize{$\phi_c$}};
  \node (captthresholdf) [above=0pt of thresholdf] {\footnotesize{$\phi_t$}};
  \draw [-,dashed] (room1) -- (controlf) -- (control) -- (thresholdf);

\caption{\label{fig:merging-graphs}A condition based on the ratio between 2
         factor graphs that share some structure can be performed by testing the
         MAP probability of a variable on a merged graph.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Detecting Novel Classes on a Graph Variable}
On the previous section it was shown how to detect novelty on a variable
by assuming $P(novel)$ is constant across any structure. It was shown that
considering that \emph{criterium} a need to compensate the likelihood of
certain graph structures for generating novel classes arises and it requires
calculating normalization factors between the structures.
That \emph{criterium} is not only a nuisance but also represents
a different problem. It can be seen as an attempt to recognize that a sample
was drawn from a novel method instead of detecting whether a given variable on
a graph is a class that the system is not aware.

This section focus on the last one. It assumes there is a distribution modelled
by a graph $G$ that is used to drawn samples. On that graph $G$ there is a
variable $a \in \{A_1,\dotsc, A_K, A^*_{K+1},\dotsc, A^*_N\}$. The system is considered
to known only the states $\{A_1,\dotsc, A_K\}$. The goal is then: given a sample subset
$x$ of variables drawn from the graph, detect whether variable
$a \in \{A_1,\dotsc, A_K\}$.

That decision can be optimally modelled by the following condition:

\begin{eqnarray*}
P_G(a \in \{A_1,\dotsc, A_K\}|x) &<& P_G(a \in \{A^*_{K+1},\dotsc,A^*_N\}|x) \\
P_G(a \in \{A_1,\dotsc, A_K\} \land x) &<& P_G(a \in \{A^*_{K+1},\dotsc,A^*_N\} \land x) \\
2 P_G(a \in \{A_1,\dotsc, A_K\} \land x) &<& P_G(x)
\end{eqnarray*}

Similarly to the previous approach a graph $G'$ can be created based on $G$ such
that conditioning on a variable $c$ allows to obtain either the conditional or
unconditional probability of a given set of variables $x$.
For that variable $a$ is extended into $a'$ by duplicating the known states $A_i$
into $A^*_i$ and extending the factors associated with it.
Additionally a control factor $\phi_c(c, a)$ is added to make $c$ enforce the
desired constraints:

\begin{eqnarray*}
a' &\in& \{A_1, \dotsc, A_K,\quad A^*_1, \dots, A^*_K,\quad A^*_{K+1}, \dots, A^*_N\} \\
c &\in& \{0, 1\}
\end{eqnarray*}
\begin{eqnarray*}
\phi_c(c, a') &=& \left\{
  \begin{array}{l l}
    1, & \qquad \text{if } c = 0 \text{ and } a' \in \{A_1,\dotsc,A_K\} \\
    1, & \qquad \text{if } c = 1 \text{ and } a' \in \{A^*_1,\dotsc,A^*_N\}\\
    0, & \qquad \text{otherwise}
  \end{array} \right. \\
\end{eqnarray*}

Note that with this transformation, the following potential equivalences hold:
\begin{eqnarray*}
\Phi_{G'}(x, c=0) &=& \Phi_{G}(x | a \in \{A_1, \dots, A_K\})\\
\Phi_{G'}(x, c=1) &=& \Phi_{G}(x) 
\end{eqnarray*}

Although those potentials allow to determine novelty of $a$, they require
knowledge of all the possible classes, which is not available.
Nonetheless access to unlabelled data or other informations or assumptions on
the unconditional distribution allow to model or approximate unconditional
distributions of $a$. For that there is interest in trying to model the set
of states $\{A^*_1, \dots, A^*_N\}$ with a set of states $\{A^{\dag}_1, \dotsc, A^{\dag}_M\}$.

Moreover assuming that the set of states $A^\dag$ is able to perfectly
characterize the distribution over any graph $G$,
for a specific graph the potentials with states $A^\dag$ are a scale-factor
away from the potentials with states $A$. And that scale-factor is dependent on
the graph structure and factors of both $G$ and $G^\dag$:

\begin{eqnarray}
P_G(x) &=& P^\dag_G(x) \\
\Phi_G(x) &=& \frac{Z_G}{Z^\dag_G} \Phi^\dag_G(x)
\end{eqnarray}

Note that this ratio, although seems similar to the one obtained in
\autoref{eq:structure-ratio} it has a special property. Here both $G$ and
$G^\dag$ model the same probability, so it is expected that both scale they
normalization factor according to some relation on how a given factor differs
from $G$ to $G^\dag$.\footnote{On \autoref{eq:structure-ratio} the ratio is
between two different distributions: a conditional and an unconditional distribution.}

If on any graph structure the marginal probabilities of all variables except $a$
is the same either by using $A$ with $\phi$, or by using $A^\dag$ with $\phi^\dag$)
then $\sum_{A^\dag}\phi^\dag_i(a^\dag, \cdot) = s_i \sum_A \phi_i(a, \cdot)$.
It can be proved that since the distribution over any graph structure has to be
equal, then the same holds for a graph structure $P_{G_i}$ composed by a single
factor $\phi_i$:

\begin{eqnarray*}
P^\dag_{G_i}(x) &=& P_{G_i}(x) \\
\frac{\phi_i(x)}{Z^\dag_{G_i}}  &=& \frac{s_i \phi_i(x)}{Z_{G_i}} \\
s_i &=& \frac{Z_{G_i}}{Z^\dag_{G_i}}
\end{eqnarray*}

With this, the scale-factor that distinguish the potential on a graph
$G$ from the potential on $G^\dag$ is given by:

\begin{eqnarray}
s &=& \frac{\Phi_G(x)}{\Phi^\dag_G(x)} \\
  &=& \frac{\sum_{a \in A}\prod \phi(a, x_\phi)}
           {\sum_{a \in A^\dag}\prod \phi^\dag(a^\dag, x_\phi)} \\
  &=& \frac{\sum_{a \in A}\prod \phi(a, x_\phi)}
           {\sum_{a \in A}\prod \phi(a, x_\phi) s_i} \\
  &=& \frac{1}{\prod s_i}
\end{eqnarray}

By adding a condition variable $c \in \{0, 1\}$, a condition factorthat is added result can be converted into a graph $G$is particularly interesting as it allows to detect novelty of $a$ by
having models to approach

Similar to the procedure followed before, a decision variable $c \in \{0, 1\}$
and a factor $\phi_c(c,a)$ will be added to $G$, creating graph $G'$.
If $\phi_c$ is defined as:

This is variable $c$ detects whether $a$ is a known or unknown class.
Note also that by adding this condition variable and factor any marginal
potential on $G$ is the same as the potential on $G'$. Since marginal potentials
can be seen as sum of potentials on all variables it can be proved by showing
the potential on any instance of all variables $X$ in $G$ is the same as the
marginal potential of the same variables in $G'$:

\begin{eqnarray*}
\Phi_{G'}(x) &=& \sum_{c}\prod_{\phi \in G'}{\phi(x_\phi)} \\
             &=& \left(\prod_{\phi \in G}{\phi(x_\phi)}\right)\phi_c(0,a) 
               + \left(\prod_{\phi \in G}{\phi(x_\phi)}\right)\phi_c(1,a)\\
             &=& \Phi_G(x)\phi_c(0,a) + \Phi_G(x)\phi_c(1,a)\\
             &=& \Phi_G(x)\\
\end{eqnarray*}

An optimal decision on whether variable $a$ of $G$ is $\in \{A_1,\dotsc,A_K\}$
can be performed by a map \gls{MAP} operation on it. And that decision can be
done solely based on potentials in $G'$:

\begin{eqnarray*}
P_G(a \in \{A_1,\dotsc,A_K\}|x) &<& P_G(a \in \{A_{K+1},\dotsc,A_N\}|x) \\
\Phi_{G'}(c = 0 \land x) &<& \Phi_{G'}(c = 1 \land x) \\
2 \Phi_{G'}(c = 0 \land x) &<& \Phi_{G'}(c=0 \land x) + \phi_{G'}(c = 1 \land x)\\
\end{eqnarray*}

By extending $a$ with duplicated states $\{A_1, \dotsc, A_K\}$, extending
the respective factors and adding a factor $\phi_f$ on the decision
variable $c$, it is possible to create a graph $G^{''}$ where a MAP decision on
$c$ will yield the same result. This graph $G^{''}$ is particullary 
the respective factors and , it is possible
to transform the previous condition on a map operation

And therefore the same happens to marginal potentials as marginal potentials
are the sum of potentials.

to detect whether a given sample from the graph has variable
$a$considered known to the system
The system only knowns $k$ of those $N$ states.
All the remaining ones are novel.

Just like in previous sections it was assumed that: it is possible to estimate
all the factors on the graph not dependent on $a$, using labelled data it is
possible to estimate the factors dependent on $a$ given that
$a \in \{A_1,\dotsc,A_K\}$. And using unlabelled data it is possible to estimate
the factors on the graph $P(x)$ for $a \in \{A_1,\dotsc,A_N\}$.

On this section it will be presented how to perform novelty detection of a
given variable class assuming samples are drawn from a distribution described
by the graph.
